The Impact of Tax Policy Evolution on Economic Development and Income Inequality

### 2024-02-20 paper/paper.qmd
Proof read the paper and fix typos on citation

“Tax Equity in Low- and Middle-Income Countries” by Pierre Bachas, Anders Jensen, and Lucie Gadenne investigates tax equity and its implications in low- and middle-income countries. They find a shift from indirect to personal income taxes as countries move towards higher levels of economic development [@article]. In low-income countries, a significantly higher percentage of revenue comes from indirect taxes (65%) compared to high-income countries (48%) [@article]. Moreover, the prevalence of the informal sector significantly affects the progressivity of indirect taxes like Value-Added Tax (VAT), contrary to common perception [@article]. The estimand focuses on these findings and explores the relationship between the transition from indirect to direct taxation and economic growth in low- and middle-income countries. We seek to quantify how changes in tax policy influence key economic indicators such as GDP growth and employment. By examining this relationship, our paper provides valuable insights into the impact of tax policy evolution on overall economic development.

@@

This paper is structured into three main sections: Data, Results, and Discussion. @sec-data Data delves into sources and methodology. It explores the various sources used for data collection, cleaning, and analyzing processes, drawing from the dataset originally published by the American Economic Association [@aea] and accessed through the openICPSR Portal [@openicpsr]. In @sec-results Results, we replicate and expand findings from the data of “Tax Equity in Low- and Middle-Income Countries” [@article], focusing on data visualizations and analyses. @sec-discussion Discussion provides a comprehensive analysis of the research findings, highlights the importance of ethical considerations in tax policy analysis, and identifies areas for future research.

@@

Data collection methods exhibit significant diversity among open sources. We conducted a data-cleaning procedure for reproduction, focusing on variable selection, simplifying variable names, and removing missing values. The variables selected from the original paper were chosen based on their relevance to our research question, relating to how does the transition from indirect to direct taxation, alongside the dynamics of the informal sector, influence tax equity and economic development in low- and middle-income countries, and what are the implications for designing effective and equitable tax policies? Throughout the process, careful attention was paid to the measurement of variables to ensure the validity and reliability of this paper’s findings. Measurement considerations included defining operational definitions for key concepts and ensuring consistency in the measurement process across different variables. The robustness check conducted by analyzing data from various years shows the data remains consistent over time.

@@

The first graph of @fig-taxrevenue-gdp-diff-source reveals a gradually declining trend in the share of tax revenue derived from indirect taxes, such as the value-added tax, sales taxes, excises, and tariffs, as a country's GDP per capita increases. This depiction underscores a noteworthy pattern wherein the reliance on indirect taxation decreases with economic advancement. The x-axis, representing GDP per capita, maintains the same log scale as in @fig-taxes-gdp, spanning from \$500 to \$50,000, while the y-axis scales from 0 to 1, indicating the proportion of total tax revenue attributed to indirect taxes. The uniformly gentle downward trajectory highlights a shifting tax structure in which nations with a lower GDP per capita have a higher propensity to lean on indirect taxes, a practice that progressively diminishes in favor of other tax forms as economies grow and develop.

@@

# Generate Figure 4 and 5
source("../scripts/100-replications-figure2.R")
```

@@

# Define the usual suspects (assuming this is a vector of variable names)
usualsuspects <- c("source", "class_pspr", "oecd", "ctry_year", "ctry_ceq", "year", "ctry", "ctry_code", "class_geo", "class_inc", "class_inc_code", "class_pspr", "class_pspr_code", "class_lend", "class_weo", "decile", "decnum", "ctry_code", "pen_scenario")

@@

# Collapse the dataset to summary statistics
# Assuming 'reportstat' is "median" and 'classificvariable' is "class_inc" as per your context
data_collapsed <- data %>%
  group_by(!!sym("class_inc"), decnum) %>%
  summarise(varofint = median(varofint, na.rm = TRUE), # Calculate median of varofint
            count = n()) %>%
  ungroup()

# Labeling variables (R doesn't use variable labels in the same way as Stata, but for documentation)
# You can store metadata or use attributes in R for a similar purpose
attr(data_collapsed$varofint, "label") <- paste("median", "dirtax", sep = "-")

# Reordering variables (dplyr automatically orders variables based on the group_by and summarise steps)
# If you need to explicitly change the order, you can use select()
data_collapsed <- data_collapsed %>%
  select(class_inc, !!sym("class_inc"), decnum, varofint, count)

@@

# Assuming 'data' is your data frame containing 'top_rate', 'gdp_pc', and 'lg_gdppc'
data <- complete_sample_data
# Create a new variable for color coding based on GDP per capita thresholds
data$color_group <- with(data, ifelse(gdp_pc >= 13000, "High GDP",
                                      ifelse(gdp_pc < 13000 & gdp_pc >= 4000, "Medium GDP", "Low GDP")))

@@

3. Challenges in Taxing High-Income Individuals: @fig-statutory-tax-rate-vs-gdp-level and @fig-workforce-share-vs-gdp-level shows the difficulties encountered in effectively taxing high-income earners in low- and middle-income countries. Issues of tax evasion and enforcement challenges are major barriers to the effectiveness of direct taxes on the wealthy. These challenges are critical for understanding the limitations of current tax systems in achieving equity, as shown in our detailed examination of tax policy implications across various economies.

A more detailed discussion on each finding will be provided in the following sections.

@@

The dynamics of tax policy are fundamentally intertwined with economic development, influencing not only fiscal stability but also social equity within nations. Taxation, encompassing both direct and indirect levies, serves as a crucial mechanism for redistributing income, funding public services, and shaping economic incentives. The relationship between tax structures and economic growth has been a subject of extensive study, revealing a nuanced interplay that varies significantly across different stages of economic development.

A pivotal aspect of this relationship is the observed shift from indirect to direct taxation as economies mature. Indirect taxes, such as sales taxes or VAT, are applied uniformly to transactions irrespective of the buyer's income level, potentially placing a disproportionate burden on lower-income households. Conversely, direct taxes, including personal and corporate income taxes, can be more effectively tailored to the payer's ability to pay, allowing for a more progressive taxation system. This shift not only reflects a country's evolving economic landscape but also its commitment to fostering social equity through tax policy.

@@

The empirical evidence from @fig-taxes-gdp and @fig-taxrevenue-gdp-diff-source in the study vividly illustrates the transition from indirect to direct taxation as economies advance. This shift is evidenced by the positive correlation between GDP per capita and the share of GDP collected through direct taxes, excluding social security contributions. The data showcases a clear trend: as countries' economies develop, there's a marked decrease in the reliance on indirect taxes such as VAT, sales taxes, and excises, with a concurrent uptick in direct taxes, particularly personal income taxes. This pattern suggests a move towards more progressive tax systems in higher GDP per capita economies, where tax policies increasingly lean on direct taxation mechanisms to foster equitable economic growth and redistribution.

This observation aligns with global trends noted in academic literature. For instance, Genschel and Seelkopf's @genschel2016tax research on taxation trends outside the OECD reveals that while non-Western countries adopt Western tax trends like increasing general revenues and implementing consumption taxes, they struggle with establishing a strong, progressive personal income tax system. This struggle is attributed to various socio-economic and structural factors, which, despite global influences, hinder the replication of progressive tax regimes seen in more developed economies.

Moreover, Stoilova and Patonov @stoilova2013empirical provide a comprehensive analysis on the impact of taxation on economic growth within the European Union, offering empirical evidence of how tax structures affect economic performance from 1995 to 2010. Their findings, alongside the examination of tax revenue trends and social contributions in European Union countries by Dubrovina et al. @dubrovina2019nexus, underscore the significance of adapting tax policies to enhance fiscal roles and ensure equitable distribution without overburdening the population.

These studies corroborate the trends depicted in the provided figures, reinforcing the global narrative of evolving tax structures as economies grow. They not only validate the empirical evidence from the study but also expand the understanding of how these trends manifest in different contexts, from OECD countries to emerging economies. Such analyses underscore the complexity and dynamism of tax policy evolution, reflecting a broad consensus on the direction of tax policy reforms aimed at fostering economic development and social equity.

@@

Literature supports this potential, with studies highlighting the role of tax policy in addressing income inequality. Ballard-Rosa, Martin, and Scheve @ballardrosa2017structure delve into the political economy aspects, suggesting that preferences over taxation significantly influence policy responses to income inequality. Moreover, Możdżierz @mozdzierz2015financialisation identifies financialization as a primary driver of increasing income inequality, indicating the need for tax policies that can counteract these trends effectively.

Further, research by Liu Qiongzhi highlights the decisive role of policy bias in determining fiscal redistribution policies, emphasizing the importance of favoring equitable tax rates and transfer payments to address income disparities [@liu2020policybias]. Khan and Padda @khan2021taxexpenditures find that in South Asia, taxation and social expenditure policies effectively reduce income inequality, highlighting the effectiveness of direct taxes and higher social spending as tools for income redistribution.

The complexity and challenges in designing effective tax policies are underscored by outliers in data, suggesting that not all shifts towards direct taxation seamlessly lead to reduced income inequality. Factors such as tax evasion and the informal economy pose significant challenges. Studies like Esaku's @esaku2021incomeinequality demonstrate how income inequality can exacerbate the size of the shadow economy, further complicating tax policy design and implementation. Additionally, Wildowicz-Szumarska @wildowiczszumarska2022eu discusses the effectiveness of redistributive policies within the EU, indicating varied success across different welfare state models in tackling income inequality through tax policy.

@@

The interplay between the informal sector and tax policy design is pivotal for economic frameworks, particularly in developing economies. @fig-informality-selfemployment and @fig-informality-consumption, showing a marked decline in informality with increased GDP per capita, underscores a critical transition phase for tax policy considerations. This visual evidence aligns with literature indicating that as economies grow, the informal sector's proportionate contribution to the economy tends to diminish, posing both challenges and opportunities for tax policy [@loayza2016informality]. Loayza's work explains how the shrinking of the informal sector as a result of economic development necessitates adaptable tax policies capable of accommodating this transition. Specifically, tax policy effectiveness hinges on recognizing the evolving tax base and ensuring that tax structures are responsive to the changing landscape of formal and informal economic activities.

The implications of these findings are profound. They suggest that policymakers must prioritize the integration of the informal sector into formal economic considerations, tailoring tax policies to be more inclusive and reflective of an economy's developmental stage. This approach not only enhances tax revenue potential but also supports the formalization of informal activities, contributing to broader economic development goals.

@@

Studies by Bird and Gendron @bird2007vat and Alm and Embaye @alm2013taxing challenge the traditional view, providing evidence that VAT and other indirect taxes can exhibit progressive characteristics in contexts where the informal sector plays a dominant role. Bird and Gendron emphasize that the perceived regressivity of VAT is not a given, suggesting that its impact varies significantly across different income groups and is influenced by the structure of the informal sector. Alm and Embaye further this argument by demonstrating that in economies where informal activities are prevalent, the consumption patterns and the tax-burden distribution across different segments of society render indirect taxes less regressive than commonly assumed.

@@

This literature underscores the importance of context-specific tax policy design, arguing for a more sophisticated analysis that considers the intricate dynamics between formal and informal economies. It points towards the potential for indirect taxes to contribute positively to tax fairness and revenue mobilization efforts in developing countries, challenging entrenched narratives about tax progressivity.

@@

Bahl and Bird @bahl2008should and Ulyssea @ulyssea2018firms offer insights into how consumption patterns within informal sectors can affect the distributional impact of indirect taxes. Bahl and Bird argue that the informal sector's consumption preferences often lean towards goods and services that are either exempt from indirect taxes or are subject to lower rates, which can inadvertently lead to a progressive tax impact. Ulyssea's work complements this view by examining the compliance aspect, suggesting that lower compliance rates in the informal sector for indirect taxes can shift the tax burden towards formal sectors, thereby influencing the overall progressivity of the tax system.

Moreover, the role of the informal sector in the broader economy, as discussed by Jaramillo and Oliva @jaramillo2013taxing, sheds light on the complex interplay between informal activities and tax policy effectiveness. They suggest that the informal sector's adaptability and its interactions with formal economic structures can mediate the impact of indirect taxes on different income groups, contributing to an unintended progressivity in tax outcomes.

@@

The datasets we used in this paper potentially generate ethical concerns and biases, particularly regarding the representation of the informal sector, marginalized populations, such as women, racial and ethnic minorities, low-income households, and individuals working in precarious employment conditions. The ethical concerns emerge from the potential exclusion or underrepresentation of these groups in the economic datasets like the World Bank’s World Development Indicators and labor market data from the International Labour Organization [@article]. This exclusion risks reinforcing systemic inequalities by formulating tax policies based on incomplete pictures of economic realities. Bird and Gendron @bird2007vat highlight the importance of considering the informal sector's impact on tax progressivity, suggesting that overlooking this aspect can lead to biased policy recommendations that fail to address the needs of a significant segment of the population.

Moreover, biases in data collection and analysis processes, as seen in the reliance on formal economic indicators, may not capture the nuanced effects of taxation on different income groups, particularly those within informal economies. The datasets may not fully account for the progressive potential of indirect taxes in environments with significant informality, potentially misrepresenting the tax burden on lower-income households. Grown and Valodia @grown2010taxation emphasize the need for tax policy analysis to include gender-disaggregated data and consider the socio-economic dynamics of informal work to mitigate biases and ensure equitable tax policy formulations.

Addressing these ethical concerns and biases needs a methodological shift towards more inclusive and comprehensive data collection and analysis strategies. Incorporating a broader spectrum of economic activities and demographic details can help capture the diverse impacts of tax policies across different segments of society. This approach aligns with Sen’s @sen1999development advocacy for equity-focused economic analysis, urging researchers to critically examine their methodologies to prevent the perpetuation of existing disparities. By embedding such considerations in tax policy research, scholars can contribute to developing more equitable and effective taxation systems that reflect the complexities of modern economies and foster inclusivity.

### Inclusivity and Representation in Tax Data
The challenge of inclusivity extends to the representation of diverse economic activities and demographic groups within tax data. Policies based on incomplete data risk perpetuating systemic inequities by failing to account for the full spectrum of economic behaviors and conditions (O'Hare, 2018). For instance, tax evasion and avoidance strategies, more accessible to affluent individuals and corporations, are not uniformly represented in tax compliance studies, potentially leading to skewed perceptions of tax fairness and effectiveness.

Ethical Research Practices
Ethical research practices demand a critical examination of these biases and the implementation of methodologies that ensure broad representation and equity in tax policy analysis. Researchers are called upon to adopt inclusive data collection methods, disaggregate data by gender, income, and other relevant demographic factors, and consciously address the systemic biases that may influence their findings (Sen, 1999).

>>>>

“Tax Equity in Low- and Middle-Income Countries” by Pierre Bachas, Anders Jensen, and Lucie Gadenne investigates tax equity and its implications in low- and middle-income countries. They find a shift from indirect to personal income taxes as countries move towards higher levels of economic development [@article]. In low-income countries, a significantly higher percentage of revenue comes from indirect taxes (65%) compared to high-income countries (48%) [@article]. Moreover, the prevalence of the informal sector significantly affects the progressivity of indirect taxes like Value-Added Tax (VAT), contrary to common perception [@article]. The estimand which we aim to replicate encompasses the aforementioned findings and explores the relationship between the transition from indirect to direct taxation and economic growth in low- and middle-income countries. We seek to quantify how changes in tax policy influence key economic indicators such as GDP growth and employment. By examining this relationship, our paper provides valuable insights into the impact of tax policy evolution on overall economic development.

@@

This paper is structured into three main sections: Data, Results, and Discussion. The Data section (@sec-data) delves into sources and methodology. It explores the various sources used for data collection, cleaning, and analyzing processes, drawing from the dataset originally published by the American Economic Association [@aea] and accessed through the openICPSR Portal [@openicpsr]. In the Results section (@sec-results), we replicate and expand findings from the data of “Tax Equity in Low- and Middle-Income Countries” [@article], focusing on data visualizations and analyses. Discussion (@sec-discussion) provides a comprehensive analysis of the research findings, highlights the importance of ethical considerations in tax policy analysis, and identifies areas for future research.

@@

Data collection methods exhibit significant diversity among open sources. To ensure the quality of our dataset, we conducted a data-cleaning procedure for reproduction, focusing on variable selection, simplifying variable names, and removing missing values. The variables selected from the original paper were chosen based on their availability and the relevance to our research question: How does the transition from indirect to direct taxation, alongside the dynamics of the informal sector, influence tax equity and economic development in low- and middle-income countries, and what are the implications for designing effective and equitable tax policies?

In the research presented within this paper, we place a strong emphasis on the detailed analysis of variable measurements to ensure the validity and reliability of our findings. This process involved defining clear operational meanings for key concepts and committing to consistent measurement across the various variables examined. By analyzing datasets from multiple years and sources, such as the World Bank's World Development Indicators and the Harvard Dataverse for oil and natural gas production, we have ensured that our data remains consistent and reliable over time. This careful approach to measurement was further supported by robustness checks, confirming the stability of our data across different periods, thus enhancing the credibility of our analyses and conclusions. Such careful attention to the measurement process highlights our commitment to conducting a comprehensive and precise exploration of the topics at hand.

@@

The first graph of @fig-taxrevenue-gdp-diff-source reveals a gradually declining trend in the share of tax revenue derived from indirect taxes, such as the value-added tax, sales taxes, excises, and tariffs, as a country's GDP per capita increases. This depiction underscores a pattern wherein the reliance on indirect taxation decreases with economic advancement. The x-axis, representing GDP per capita, maintains the same log scale as in @fig-taxes-gdp, spanning from \$500 to \$50,000, while the y-axis scales from 0 to 1, indicating the proportion of total tax revenue attributed to indirect taxes. The uniformly gentle downward trajectory highlights a shifting tax structure in which nations with a lower GDP per capita have a higher propensity to lean on indirect taxes, a practice that progressively diminishes in favor of other tax forms as economies grow and develop.

@@

# Generate Figure 3 and 4, stored in plot1 and plot2
source("../scripts/100-replications-figure2.R")
```

@@

# Define the usual suspects
usualsuspects <- c("source", "class_pspr", "oecd", "ctry_year", "ctry_ceq", "year", "ctry", "ctry_code", "class_geo", "class_inc", "class_inc_code", "class_pspr", "class_pspr_code", "class_lend", "class_weo", "decile", "decnum", "ctry_code", "pen_scenario")

@@

# Collapse the dataset to summary statistics
data_collapsed <- data %>%
  group_by(!!sym("class_inc"), decnum) %>%
  summarise(varofint = median(varofint, na.rm = TRUE), # Calculate median of varofint
            count = n()) %>%
  ungroup()

# Labeling variables
attr(data_collapsed$varofint, "label") <- paste("median", "dirtax", sep = "-")

# Reordering variables
data_collapsed <- data_collapsed %>%
  select(class_inc, !!sym("class_inc"), decnum, varofint, count)

@@

data <- complete_sample_data
data$color_group <- with(data, ifelse(gdp_pc >= 13000, "High GDP",
                                      ifelse(gdp_pc < 13000 & gdp_pc >= 4000, "Medium GDP", "Low GDP")))

@@

A comprehensive analysis of each finding will be elaborated upon in the subsequent sections.

@@

The relationship between tax structures and economic growth has been a subject of extensive study. A pivotal aspect of this relationship is the observed shift from indirect to direct taxation as economies mature. Indirect taxes, such as sales taxes or VAT, are applied uniformly to transactions irrespective of the buyer's income level, potentially placing a disproportionate burden on lower-income households. Conversely, direct taxes, including personal and corporate income taxes, can be more effectively tailored to the payer's ability to pay, allowing for a more progressive taxation system. This shift not only reflects a country's evolving economic landscape but also its commitment to fostering social equity through tax policy.

@@

The empirical evidence from @fig-taxes-gdp and @fig-taxrevenue-gdp-diff-source in the study illustrates the transition from indirect to direct taxation as economies advance. This shift is evidenced by the positive correlation between GDP per capita and the share of GDP collected through direct taxes, excluding social security contributions. The data showcases a clear trend: as countries' economies develop, there's a marked decrease in the reliance on indirect taxes such as VAT, sales taxes, and excises, with a concurrent uptick in direct taxes, particularly personal income taxes. This pattern suggests a move towards more progressive tax systems in higher GDP per capita economies, where tax policies increasingly lean on direct taxation mechanisms to foster equitable economic growth and redistribution.

This observation aligns with global trends noted in academic literature. For instance, @genschel2016tax conducted a research on taxation trends outside the OECD and revealed that while non-Western countries adopt Western tax trends like increasing general revenues and implementing consumption taxes, they struggle with establishing a strong, progressive personal income tax system. This struggle is attributed to various socio-economic and structural factors, which, despite global influences, hinder the replication of progressive tax regimes seen in more developed economies.

Moreover, @stoilova2013empirical provide a comprehensive analysis on the impact of taxation on economic growth within the European Union, offering empirical evidence of how tax structures affect economic performance from 1995 to 2010. Their findings, alongside the examination of tax revenue trends and social contributions in European Union countries by @dubrovina2019nexus, underscore the significance of adapting tax policies to enhance fiscal roles and ensure equitable distribution without overburdening the population.

These studies corroborate the trends shown in the provided figures, reinforcing the global narrative of evolving tax structures as economies grow. They not only validate the empirical evidence from the study but also expand the understanding of how these trends manifest in different contexts, from OECD countries to emerging economies. Such analyses underscore the complexity and dynamism of tax policy evolution, reflecting a broad consensus on the direction of tax policy reforms aimed at fostering economic development and social equity.

@@

Literature supports this potential, with studies highlighting the role of tax policy in addressing income inequality. @ballardrosa2017structure delve into the political economy aspects, suggesting that preferences over taxation significantly influence policy responses to income inequality. Moreover, @mozdzierz2015financialisation identifies financialization as a primary driver of increasing income inequality, indicating the need for tax policies that can counteract these trends effectively.

Further, research by Liu Qiongzhi highlights the decisive role of policy bias in determining fiscal redistribution policies, emphasizing the importance of favoring equitable tax rates and transfer payments to address income disparities [@liu2020policybias]. @khan2021taxexpenditures find that in South Asia, taxation and social expenditure policies effectively reduce income inequality, highlighting the effectiveness of direct taxes and higher social spending as tools for income redistribution.

The complexity and challenges in designing effective tax policies are underscored by outliers in data, suggesting that not all shifts towards direct taxation seamlessly lead to reduced income inequality. Factors such as tax evasion and the informal economy pose significant challenges. Studies like Esaku's demonstrate how income inequality can exacerbate the size of the shadow economy, further complicating tax policy design and implementation [@esaku2021incomeinequality]. Additionally, @wildowiczszumarska2022eu discusses the effectiveness of redistributive policies within the EU, indicating varied success across different welfare state models in tackling income inequality through tax policy.

@@

The interplay between the informal sector and tax policy design is pivotal for economic frameworks, particularly in developing economies. @fig-informality-selfemployment and @fig-informality-consumption, showing a decline in informality with increased GDP per capita, underscores a critical transition phase for tax policy considerations. This visual evidence aligns with literature indicating that as economies grow, the informal sector's proportionate contribution to the economy tends to diminish, posing both challenges and opportunities for tax policy [@loayza2016informality]. Loayza's work explains how the shrinking of the informal sector as a result of economic development necessitates adaptable tax policies capable of accommodating this transition. Specifically, tax policy effectiveness hinges on recognizing the evolving tax base and ensuring that tax structures are responsive to the changing landscape of formal and informal economic activities.

The implications of these findings suggest that policymakers must prioritize the integration of the informal sector into formal economic considerations, tailoring tax policies to be more inclusive and reflective of an economy's developmental stage. This approach not only enhances tax revenue potential but also supports the formalization of informal activities, contributing to broader economic development goals.

@@

Studies by @bird2007vat and @alm2013taxing challenge the traditional view, providing evidence that VAT and other indirect taxes can exhibit progressive characteristics in contexts where the informal sector plays a dominant role. Bird and Gendron emphasize that the perceived regressivity of VAT is not a given, suggesting that its impact varies significantly across different income groups and is influenced by the structure of the informal sector. @alm2013taxing further this argument by demonstrating that in economies where informal activities are prevalent, the consumption patterns and the tax-burden distribution across different segments of society render indirect taxes less regressive than commonly assumed.

@@

This paper underscores the importance of context-specific tax policy design, arguing for a more sophisticated analysis that considers the intricate dynamics between formal and informal economies. It points towards the potential for indirect taxes to contribute positively to tax fairness and revenue mobilization efforts in developing countries, challenging entrenched narratives about tax progressivity.

@@

@bahl2008should and @ulyssea2018firms offer insights into how consumption patterns within informal sectors can affect the distributional impact of indirect taxes. Bahl and Bird argue that the informal sector's consumption preferences often lean towards goods and services that are either exempt from indirect taxes or are subject to lower rates, which can inadvertently lead to a progressive tax impact. Ulyssea's work complements this view by examining the compliance aspect, suggesting that lower compliance rates in the informal sector for indirect taxes can shift the tax burden towards formal sectors, thereby influencing the overall progressivity of the tax system.

Moreover, the role of the informal sector in the broader economy, as discussed by @jaramillo2013taxing, sheds light on the complex interplay between informal activities and tax policy effectiveness. They suggest that the informal sector's adaptability and its interactions with formal economic structures can mediate the impact of indirect taxes on different income groups, contributing to an unintended progressivity in tax outcomes.

@@

The datasets we used in this paper potentially generate ethical concerns and biases, particularly regarding the representation of the informal sector, marginalized populations, such as women, racial and ethnic minorities, low-income households, and individuals working in precarious employment conditions. The ethical concerns emerge from the potential exclusion or underrepresentation of these groups in the economic datasets like the World Bank’s World Development Indicators and labor market data from the International Labour Organization [@article]. This exclusion risks reinforcing systemic inequalities by formulating tax policies based on incomplete pictures of economic realities. @bird2007vat highlight the importance of considering the informal sector's impact on tax progressivity, suggesting that overlooking this aspect can lead to biased policy recommendations that fail to address the needs of a significant segment of the population.

Moreover, biases in data collection and analysis processes, as seen in the reliance on formal economic indicators, may not capture the nuanced effects of taxation on different income groups, particularly those within informal economies. The datasets may not fully account for the progressive potential of indirect taxes in environments with significant informality, potentially misrepresenting the tax burden on lower-income households. @grown2010taxation emphasize the need for tax policy analysis to include gender-disaggregated data and consider the socio-economic dynamics of informal work to mitigate biases and ensure equitable tax policy formulations.

Addressing these ethical concerns and biases needs a methodological shift towards more inclusive and comprehensive data collection and analysis strategies. Incorporating a broader spectrum of economic activities and demographic details can help capture the diverse impacts of tax policies across different segments of society. This approach aligns with @sen1999development's advocacy for equity-focused economic analysis, urging researchers to critically examine their methodologies to prevent the perpetuation of existing disparities. By embedding such considerations in tax policy research, scholars can contribute to developing more equitable and effective taxation systems that reflect the complexities of modern economies and foster inclusivity.

### Inclusivity and Representation in Tax Data
The challenge of inclusivity extends to the representation of diverse economic activities and demographic groups within tax data. Policies based on incomplete data risk perpetuating systemic inequities by failing to account for the full spectrum of economic behaviors and conditions [@ohare2018difficult]. For instance, tax evasion and avoidance strategies, more accessible to affluent individuals and corporations, are not uniformly represented in tax compliance studies, potentially leading to skewed perceptions of tax fairness and effectiveness.

Ethical Research Practices
Ethical research practices demand a critical examination of these biases and the implementation of methodologies that ensure broad representation and equity in tax policy analysis. Researchers are called upon to adopt inclusive data collection methods, disaggregate data by gender, income, and other relevant demographic factors, and consciously address the systemic biases that may influence their findings [@sen1999development].
